---
title: "Forcasting US Presidential Election of 2024"
subtitle: "Donald Trump Is Expected To Win Popular Vote By Linear Model"  
author:
  - Yuxin Sun
thanks: "A GitHub Repository containing all data, R code, and other files used in this investigation is located here: <https://github.com/Yuxin-Sun-Caroline/US_Presidential_Forecast_2024.git>"
date: 10-22-2024
date-format: long
abstract: "The 2024 U.S. Presidential Election is shaping up to be a rematch between former President Donald Trump and Vice President Kamala Harris, following President Joe Biden’s withdrawal from the race in mid-2024. In this analysis, we employ a linear regression model to project the election outcome. Current model results indicate a tight race, with Trump holding a narrow lead in the Electoral College. However, high polling volatility and potential 'October surprises' emphasize uncertainties. Future research should explore the evolving dynamics in voter sentiment and potential implications for policy."
format: pdf
number-sections: true
# The toc parameter used to create a table of contents was obtained from
# Quarto documentation:
# https://quarto.org/docs/reference/formats/pdf.html
toc: true
bibliography: references.bib
---

```{r}
#| echo: false
#| warning: false
#### Preamble ####

# Purpose: Generates a paper titled "Forcasting US Presidential Election of 2024" in PDF format when rendered.
# Author: Yuxin Sun
# Date:   Octomber 22, 2024
# Contact: yuxin.sun@mail.utoronto.ca
# Pre-requisites: none".
```

```{r}
#| echo: false
#| warning: false
#### Loading Packages ####

# install.packages("tidyverse")
# install.packages("here")
library(tidyverse)
library(here)
```

# Introduction {#sec-introduction}
The 2024 U.S. Presidential Election, set for November 5th, presents a pivotal rematch between former President Donald Trump and Vice President Kamala Harris. With Biden’s withdrawal and Trump’s lead among Republican delegates, this election symbolizes a deeply divided political landscape. Amid record-low trust in institutions and heightened polarization, understanding potential outcomes is crucial. For this report, we employ a linear model with multilevel post-stratification (MRP) to predict the popular vote and Electoral College results. Using key demographic and survey data, we examine voter dynamics and outline the implications for American democracy.

# Data {#sec-data}

## Overview {#sec-data-overview}
For this analysis, the R programming language [@citeR] was used along with packages such as *dplyr* [@citedplyr] and *knitr* [@citeknitr] to clean and organize the data. The *here* [@citehere] was used to find reference files. The *opendatatoronto* package [@citeopendatatoronto] was utilized to download the raw data directly from Toronto’s Open Data Portal, followed by data cleaning and analysis using the *tidyverse*[@citetidyverse] suite of packages. The *ggplot2*[@citeggplot2], *RColorBrewer*[@citeRcolorbrewer],*scales*[@citescales] and *dplyr*[@citedplyr] to visualize the graphs. This process ensured the accuracy and reliability of the analysis, allowing for insights into the trends within Toronto’s shelter system over time.

## Raw data
The dataset used in this analysis is part of the "Toronto Shelter System Flow Data[@citeODTshelter]," which is regularly updated and published by the City of Toronto[@citeopendatatoronto]. This dataset provides comprehensive details on the shelter system's capacity, usage, and flow of individuals across different shelters in the city. 
The dataset captures various variables, including the number of actively homeless individuals, categorized by demographic groups such as gender, age, and refugee status. Key variables analyzed in this report include “Shelter Capacity” and “Occupancy Rate,” which track the shelter system's capacity and how much of it is being used, respectively, as well as “Actively Homeless Population,” a variable reflecting the number of individuals experiencing homelessness. This data is crucial for understanding the demographic shifts in Toronto's homeless population and the system’s ability to accommodate these changes.

## Cleaning data
In the process of cleaning the homeless dataset, several adjustments were made to ensure accuracy and facilitate analysis. First, all instances of percentage values were stored as strings with a "%" symbol, making them non-numeric. These entries were cleaned by removing the "%" symbol and converting the values to numeric form, allowing for proper quantitative analysis. Additionally, the original dataset contained a number of columns with complex or inconsistent names, such as `age16.24` and `gender_transgender.non.binary_or_two_spirit`, which were standardized and renamed to more intuitive labels like `age_16_24` and `gender_transgender_non_binary_or_two_spirit`, respectively. This improved clarity and streamlined the data manipulation process. Furthermore, all unnecessary columns that were not directly related to the analysis were retained but kept consistent for future exploration. 



```{r}
#| echo: false
#| warning: false
#| message: false

#### Clean the Raw Dataset and Label for the graph ####

# Load necessary library
library(dplyr)

# Load the data
data <- read.csv("/Users/so_watermelon/Downloads/starter_folder-main-2/data/raw_data/raw_data.csv")

# Clean column names
clean_data <- data %>%
  rename(
    id = X_id,
    date = date.mmm.yy.,
    population_group = population_group,
    returned_from_housing = returned_from_housing,
    returned_to_shelter = returned_to_shelter,
    newly_identified = newly_identified,
    moved_to_housing = moved_to_housing,
    became_inactive = became_inactive,
    actively_homeless = actively_homeless,
    age_under_16 = ageunder16,
    age_16_24 = age16.24,
    age_25_34 = age25.34,
    age_35_44 = age35.44,
    age_45_54 = age45.54,
    age_55_64 = age55.64,
    age_65_over = age65over,
    gender_male = gender_male,
    gender_female = gender_female,
    gender_transgender_non_binary_or_two_spirit = gender_transgender.non.binary_or_two_spirit,
    population_group_percentage = population_group_percentage
  )

# Convert percentage column to numeric by removing the percentage sign
clean_data <- clean_data %>%
  mutate(population_group_percentage = as.numeric(gsub("%", "", population_group_percentage)))

# Display the cleaned data
#head(clean_data)
library(knitr)
#| echo: false
#| warning: false
#| label: tbl-cleaned-homeless-data
#| tbl-cap: Sample of cleaned homeless data 
#| echo: false

head(clean_data, 6) |>
  kable(
    col.names = c("ID", "Date", "Population Group", "Returned From Housing", "Returned to Shelter", 
                  "Newly Identified", "Moved to Housing", "Became Inactive", "Actively Homeless", 
                  "Age <16", "Age 16-24", "Age 25-34", "Age 35-44", "Age 45-54", "Age 55-64", 
                  "Age 65+", "Male", "Female", "Trans/Non-binary/Two-Spirit", "Pop Group Percentage"),
    booktabs = TRUE,
    caption = '<strong style="font-size: 19px;">Sample of cleaned homeless data</strong>'
  )
```

A sample of cleaned data is represented above with Table 1.
After loading the dataset using the R programming language [@citeR] and the `here` package [@citehere], the `tidyverse` [@citetidyverse] package was used to generate graphs. In doing so, R code was adapted from @tellingstorieswithdata.
For each of the categories in this report, we have specific definitions to represent the different groups involved. **All populations** refers to the total number of individuals without demographic breakdown, offering an overall count. **Chronic homelessness** is represented by people who meet federal criteria, either by having stayed in shelters for 180 nights in the past year or for a total of at least 546 nights over the past three years. **Families** represent individuals staying in family-designated services, while **youth** refers to unaccompanied individuals aged 16 to 24 who are not part of a family. **Single adults** are those who do not fall under youth or family classifications. **Refugees** are individuals identified as such during shelter intake or through specific programs, while **non-refugees** are those who do not meet the refugee criteria. Finally, **Indigenous** refers to individuals who have self-identified as First Nations, Métis, or Inuit during at least one intake recorded in the Shelter Management Information System (SMIS). These categories help clarify distinctions and target interventions more effectively.
\newpage

## Results {#sec-data-results}

```{r}
#| echo: false
#| warning: false
#| label: fig-bar-chart-of-distribution
#| fig-cap: "Actively Homeless Population By Group from 2018 to 2024 in Toronto"

#### First Graph ####

# Install and load necessary libraries if not already installed
if (!require(ggplot2)) install.packages("ggplot2", dependencies=TRUE)
if (!require(RColorBrewer)) install.packages("RColorBrewer", dependencies=TRUE)
library(ggplot2)
library(RColorBrewer)

# Define a custom rainbow-like Morandi-inspired color palette with a better balance
morandi_rainbow_palette_balanced <- c("#D4A8C0", "#A8D4E0", "#C0A8D4", "#D4A8A8", 
                                      "#F4B183", "#E0A8D4", "#A8C0D4", "#D4E0A8")

# Bar plot of actively homeless population by population group with the updated balanced Morandi rainbow colors
ggplot(clean_data, aes(x = population_group, y = actively_homeless, fill = population_group)) +
  geom_bar(stat = "identity", alpha = 0.9) +  # Bar plot without black borders
  scale_fill_manual(values = morandi_rainbow_palette_balanced) +         # Apply the updated rainbow-like palette
  labs(title = "Actively Homeless Population by Group",
       x = "Population Group",
       y = "Actively Homeless Population",
       fill = "Population Group") +
  theme_minimal(base_size = 15) +                                         # Minimal theme for a clean look
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),     # Center the title and make it bold
    axis.text.x = element_text(angle = 45, hjust = 1),                    # Rotate x-axis labels for readability
    legend.position = "none",                                             # Remove the legend for simplicity
    panel.grid.major = element_line(color = "#E5E5E5")                    # Soft gridlines for a subtle look
  )


```

@fig-bar-chart-of-distribution clearly illustrates the comparison and differences in the number of individuals identified as "actively homeless" across different groups in Toronto shelter data base. The "all population" category, representing the majority without specific demographic breakdown, shows a strikingly high number of homeless individuals. When comparing families with single adults, it is evident that single adults significantly outnumber families. Similarly, when comparing refugees with non-refugees, non-refugees far exceed the number of refugees.




```{r}
#| echo: false
#| warning: false
#| label: fig-gender-comparison
#| fig-cap: "Porportion of different genders in Actively Homeless Population"

#### Second Graph ####

# Install and load necessary libraries if not already installed
if (!require(ggplot2)) install.packages("ggplot2", dependencies=TRUE)
if (!require(dplyr)) install.packages("dplyr", dependencies=TRUE)
if (!require(scales)) install.packages("scales", dependencies=TRUE)  # For percentage labels
library(ggplot2)
library(dplyr)
library(scales)

# Define colors for gender
gender_colors <- c("male" = "#A0E7E5",       
                   "female" = "#D4A5A5",     
                   "transgender" = "#9E9E9E") 

# Summarize the total counts by gender
gender_data <- clean_data %>%
  summarise(
    male = sum(gender_male, na.rm = TRUE),
    female = sum(gender_female, na.rm = TRUE),
    transgender = sum(gender_transgender_non_binary_or_two_spirit, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = everything(), names_to = "gender", values_to = "count")

# Add percentage labels
gender_data <- gender_data %>%
  mutate(percentage = count / sum(count) * 100,
         label = paste0(gender, " (", round(percentage, 1), "%)"))

# Create the pie chart
ggplot(gender_data, aes(x = "", y = count, fill = gender)) +
  geom_bar(stat = "identity", width = 1, color = "white") +  # Create the bar for pie chart
  coord_polar(theta = "y") +  # Convert bar chart to pie chart
  scale_fill_manual(values = gender_colors, 
                    labels = c("Male", "Female", "Transgender/Non-binary")) +  # Apply custom colors
  geom_text(aes(label = label), position = position_stack(vjust = 0.5), color = "white", size = 5) +  # Add percentage labels
  labs(title = "Gender Distribution in Population") +
  theme_void() +  # Remove axes and grid for clean pie chart
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),  # Center and bold the title
    legend.title = element_blank(),                                    # Remove legend title
    legend.position = "right"                                          # Position the legend on the right
  )



```

@fig-gender-comparison shows that the vast majority of actively homeless in Toronto were males (approximately 61.6\%). Females on the other hand, shows a lot less percentage by (37\%), which makes approcimately nearly a half less than the males. For Transgender or non-binary group, they played only (1.4\%) in this distribution. However, such a low proportion could also be due to the fact that transgender individuals may not openly disclose their gender.


```{r}
#| echo: false
#| warning: false
#| label: fig-line-graph-by-time
#| fig-cap: "Active Homeless Population Change From 2018 to 2024"
#### Third Graph ####


# Load necessary libraries
library(ggplot2)
library(dplyr)

# Assuming the cleaned dataset is already loaded into a data frame called 'clean_data'

# Convert the 'date' column to a date format for plotting
clean_data <- clean_data %>%
  mutate(date = as.Date(paste("01", date, sep = "-"), format = "%d-%b-%y"))

# Create a line plot of active homeless population over time
ggplot(clean_data, aes(x = date, y = actively_homeless, group = population_group, color = population_group)) +
  geom_line(size = 1) +
  labs(title = "Active Homeless Population Over Time",
       x = "Date",
       y = "Number of Actively Homeless",
       color = "Population Group") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

@fig-line-graph-by-time illustrates the growth history and trends of different categories within the actively homeless population from 2018 to 2024. Notably, starting from 2022, there has been a significant increase in chronic, refugee, and single adult populations, in addition to the overall "all population" category. The year 2022, marked by the initial economic impacts of the COVID-19 pandemic and emerging global geopolitical challenges, highlights the severe toll these shifts have taken on vulnerable groups, including single adults and refugees. The rise in the chronic homeless population further indicates that the long-term effects of these challenges will prolong the difficulty and pressure on efforts to restore safety and stability in communities.



# Discussion {#sec-discussion}

In @fig-bar-chart-of-distribution, the comparison between different groups within Toronto’s shelter system clearly highlights significant disparities in the actively homeless population. The all population category, representing the majority, shows a strikingly large number of homeless individuals without demographic breakdowns. When focusing on specific groups, single adults significantly outnumber families, and non-refugees far exceed refugees. These findings align with the broader trends observed in @fig-line-graph-by-time, which shows a marked increase in chronic, refugee, and single adult homeless populations beginning in 2022, likely driven by the economic and social effects of the COVID-19 pandemic. This sharp rise in homelessness within these vulnerable groups underscores the ongoing and prolonged impacts of these global crises, making it more difficult to restore community stability.

Gender distribution data from @fig-gender-comparison further reveals that males constitute the vast majority of Toronto’s homeless population at 61.6%, with females making up 37%. The proportion of transgender or non-binary individuals is notably low at just 1.4%, though this could be partly due to underreporting, as some individuals may not feel comfortable disclosing their gender identity. These combined findings emphasize the urgent need for targeted policies that address the distinct needs of these subgroups, especially given the increasing challenges observed in recent years.

# Limitations and future {#sec-limit-future}
There are several limitations to the analysis conducted and the conclusions drawn regarding Toronto’s homeless population. Firstly, the dataset only reflects individuals who have utilized overnight shelter services, excluding those exclusively sleeping outdoors or using alternative homelessness services. As a result, this may lead to an underestimation of the total homeless population, as a significant segment of individuals experiencing homelessness may not be represented. Furthermore, shelter sites that do not utilize the Shelter Management Information System (SMIS) and are funded by other levels of government are also not included in the dataset, creating additional gaps in the analysis.

Another important limitation is that the Open Data set is updated monthly, typically on the 15th day of the month. The reason for this delay is to allow the discharge field in SMIS to remain active for two weeks, enhancing the accuracy of the final discharge records. While this system increases accuracy, it also means that the dataset for any given month is incomplete until the following update is published. Thus, the dataset is in a constant state of revision, and previous month data are replaced when a new extraction is complete, which could affect the consistency of conclusions over time.

Additionally, while extensive data on Toronto’s shelter system flow from 2018 to 2024 was analyzed, demographic information such as gender and refugee status relies on self-reporting. This reliance on self-reported data may result in underreporting, especially among groups like transgender or non-binary individuals who may not feel comfortable disclosing their identity.

Future research should aim to include a broader spectrum of the homeless population, particularly those not using formal shelter services, to gain a more comprehensive understanding of homelessness in Toronto. Moreover, expanding the analysis to explore the long-term effects of factors like the COVID-19 pandemic on the chronic homeless population, as well as housing stability and service utilization across other Canadian cities, could provide deeper insights and inform more effective interventions. In addition to these efforts, government projects should prioritize targeted support for specific populations, including refugees, single adults, and men, who represent a significant portion of long-term homelessness. By focusing on these vulnerable groups, the government could help reduce the strain on shelters, creating more capacity and improving overall system efficiency. Addressing the unique challenges faced by these populations could not only enhance their quality of life but also alleviate pressure on the city’s shelter system, allowing for more sustainable solutions.


\newpage

\appendix

# Appendix {#sec-appendix}

## Dataset and Graph Sketches {#sec-appendix-sketches}
This paper analyzes the homeless population in Toronto using data from the Shelter System Flow between 2018 and 2024, with a focus on identifying the major target groups within the actively homeless population. By utilizing a sample of 604 individuals, the research examines the distribution of key demographic categories, such as 'single adult,' 'non-refugee,' and 'male.' The findings reveal that these categories make up a significant portion of the actively homeless population, with a marked growth trend in the 'single adult' and 'refugee' groups. This analysis underscores the urgent need for targeted policies and interventions to address the increasing demand for shelter services, particularly for single adults and refugees, in order to reduce homelessness and improve the efficiency of the city’s support systems.


## Data Cleaning {#sec-appendix-cleaning}
To clean the data, we first load the raw CSV file using the `read.csv` function in R, ensuring that the data is read correctly into a data frame. The next step involves cleaning and standardizing the column names, which may contain dots or unconventional symbols. By using the `rename()` function from the `dplyr` package, we give the columns more descriptive and consistent names, improving readability and facilitating future analysis. After that, we handle the `population_group_percentage` column, which contains percentages in string format (e.g., "100%"). We remove the percentage sign and convert the values to numeric using `gsub()` and `as.numeric()` to make them suitable for analysis.

If the date column is in a string format, we also convert it into a proper date format using `mutate()` and `as.Date()`, which ensures that the dates can be used for plotting time series data. After these steps, we check the structure of the cleaned data using `str()` to verify that all the transformations are correctly applied, ensuring the data is ready for further analysis and visualization.

This process cleans and organizes the data, making it easier to perform insightful analysis and generate meaningful visualizations, such as line plots and bar graphs.


## Attribution Statement {#sec-appendix-attribution}

"Contains information licensed under the Open Government Licence – Toronto" [@citeLicense].

\newpage

# References {#sec-references}